---
title: "STAT6011 Assignment 2"
author: "Lee Lap San 3036373312"
date: "2025-11-02"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

### a

```{r 1a1}
# Load necessary libraries
library(ggplot2)
library(cluster)

# Load the dataset
data <- read.csv("q1.csv")

# Assuming the data is in a column named 'x'
x <- data$y

## a
# 1. Visualize the data using a histogram
ggplot(data, aes(x = x)) +
  geom_histogram(binwidth = 0.15, fill = "skyblue", color = "black") +
  labs(title = "Histogram of Observed Data", x = "x", y = "Frequency")

```
As shown in the histogram, there appears to be four peaks. We make a guess that there are 4 clusters. We estimate $\{\mu_k, \sigma_k\}_{k=1}^4$ by sample statistics.
```{r 1a2}
# 2. Apply hierarchical clustering
# First, reshape the data for clustering
x_matrix <- matrix(x, ncol = 1)

# Compute the distance matrix
dist_matrix <- dist(x_matrix)

# Perform hierarchical clustering
hc <- hclust(dist_matrix, method = "ward.D2")

# Plot the dendrogram
plot(hc, labels = FALSE, main = "Hierarchical Clustering Dendrogram")

# Cut the tree to guess K (e.g., try 2 to 5 clusters)
K_guess <- 4  # Guess
clusters <- cutree(hc, k = K_guess)

# Estimate parameters for each cluster
for (k in 1:K_guess) {
  cluster_data <- x[clusters == k]
  mu_k <- mean(cluster_data)
  sigma_k <- sd(cluster_data)
  cat(sprintf("Cluster %d: Mean = %.2f, SD = %.2f\n", k, mu_k, sigma_k))
}
```

### b
We consider a Bayesian Gaussian mixture model with a number of components $K$. Each observation $x_i$ is modeled as:
\begin{equation}
x_i \sim \sum_{k=1}^K \pi_k \mathcal{N}(\mu_k, \sigma_k^2), \quad i = 1, \dots, n
\end{equation}

where $\pi_k$ are the mixing proportions such that $\sum_{k=1}^K \pi_k = 1$, and $\mu_k, \sigma_k^2$ are the mean and variance of the $k$-th component.
We introduce latent allocation variables $z_i \in \{1, \dots, K\}$ such that:

\begin{align}
z_i \sim \text{Categorical}(\pi_1, \dots, \pi_K) \\
x_i \mid z_i = k \sim \mathcal{N}(\mu_k, \sigma_k^2) \\
\end{align}

The complete-data likelihood is:
\begin{align}
L(\{\mu_k,\sigma_k,\pi_k\}_{k=1}^K \mid \mathbf{x},\mathbf{z}) \\
= \prod_{i=1}^n[\pi_{z_i}\frac{1}{\sigma_{z_i}}exp(-\frac{1}{2}(\frac{x_i-\mu_{z_i}}{\sigma_{z_i}})^2)]
\end{align}

The prior distributions are:
\begin{align}
\mu_k \sim \mathcal{N}(0, 1) \\
\tau_k = \sigma_k^{-2} \sim \text{Gamma}(0.1, 0.1) \\
(\pi_1, \dots, \pi_K) \sim \text{Dirichlet}(1, \dots, 1) \\
\end{align}


Complete Conditional Distributions
These arise from the conjugacy between the priors and the likelihoods
Conditional for $z_i$:
\begin{equation*}
p(z_i = k \mid x_i, \mu_k, \sigma_k^2, \pi_k) \propto \pi_k \cdot \mathcal{N}(x_i \mid \mu_k, \sigma_k^2)
\end{equation*}

Conditional for $\mu_k$:
Given the normal prior and normal likelihood:
\begin{equation*}
\mu_k \mid \{x_i : z_i = k\}, \tau_k \sim \mathcal{N}\left( \frac{\tau_k \sum_{i: z_i = k} x_i}{\tau_k n_k + 1}, \frac{1}{\tau_k n_k + 1} \right)
\end{equation*}

Conditional for $\tau_k$:
Given the Gamma prior and normal likelihood:
\begin{equation*}
\tau_k \mid \{x_i : z_i = k\}, \mu_k \sim \text{Gamma}\left( a + \frac{n_k}{2}, \; b + \frac{1}{2} \sum_{i: z_i = k} (x_i - \mu_k)^2 \right)
\end{equation*}

Conditional for $\pi$:
Given the Dirichlet prior and categorical likelihood:
\begin{align}
\pi \mid \{z_i\} \sim \text{Dirichlet}(\alpha_1 + n_1, \dots, \alpha_K + n_K) \\
n_k = \sum_{i=1}^N I(z_i=k)
\end{align}


### c
```{r 1c}
## c
n <- length(x)
K <- 3
sigma2 <- rep(1, K)
pi_k <- rep(1/K, K)

# Initialize parameters
mu <- rnorm(K, 0, 1)
z <- sample(1:K, n, replace = TRUE)

# Gibbs sampling settings
n_iter <- 10000
burn_in <- 1000
mu_samples <- matrix(NA, nrow = n_iter, ncol = K)

# Gibbs sampler
for (iter in 1:n_iter) {
  # Step 1: Sample z_i
  for (i in 1:n) {
    probs <- sapply(1:K, function(k) {
      dnorm(x[i], mean = mu[k], sd = sqrt(sigma2[k])) * pi_k[k]
    })
    z[i] <- sample(1:K, 1, prob = probs / sum(probs))
  }
  
  # Step 2: Sample mu_k
  for (k in 1:K) {
    x_k <- x[z == k]
    n_k <- length(x_k)
    if (n_k > 0) {
      x_bar_k <- mean(x_k)
      post_var <- 1 / (n_k + 1)
      post_mean <- n_k * x_bar_k * post_var
      mu[k] <- rnorm(1, mean = post_mean, sd = sqrt(post_var))
    } else {
      mu[k] <- rnorm(1, 0, 1)  # fallback if no data assigned
    }
  }
  
  mu_samples[iter, ] <- mu
}

# Remove burn-in
mu_samples_post <- mu_samples[(burn_in + 1):n_iter, ]

# Posterior summaries
posterior_means <- colMeans(mu_samples_post)
posterior_vars <- apply(mu_samples_post, 2, var)

cat("Posterior Means:\n")
print(posterior_means)
cat("Posterior Variances:\n")
print(posterior_vars)

# Trace plots
par(mfrow = c(1, 3))
for (k in 1:K) {
  plot(mu_samples_post[, k], type = "l", main = paste("Trace Plot for mu", k),
       xlab = "Iteration", ylab = expression(mu[k]))
}
```

### d
```{r 1d}
# Load necessary libraries
library(ggplot2)
library(MCMCpack)  # for Dirichlet sampling

n <- length(x)
K <- 3

# Priors
mu_prior_mean <- 0
mu_prior_var <- 1
tau_shape <- 0.1
tau_rate <- 0.1
alpha_pi <- rep(1, K)

# Initialization
set.seed(123)
z <- sample(1:K, n, replace = TRUE)
mu <- rnorm(K, mu_prior_mean, sqrt(mu_prior_var))
tau <- rgamma(K, tau_shape, rate = tau_rate)
pi <- rdirichlet(1, alpha_pi)[1, ]

# Storage
n_iter <- 10000
burn_in <- 1000
mu_samples <- matrix(NA, nrow = n_iter, ncol = K)
tau_samples <- matrix(NA, nrow = n_iter, ncol = K)
pi_samples <- matrix(NA, nrow = n_iter, ncol = K)

# Gibbs sampler
for (iter in 1:n_iter) {
  # Step 1: Sample z_i
  for (i in 1:n) {
    probs <- sapply(1:K, function(k) {
      pi[k] * dnorm(x[i], mean = mu[k], sd = sqrt(1 / tau[k]))
    })
    z[i] <- sample(1:K, 1, prob = probs / sum(probs))
  }
  
  # Step 2: Sample mu_k
  for (k in 1:K) {
    x_k <- x[z == k]
    n_k <- length(x_k)
    if (n_k > 0) {
      post_var <- 1 / (tau[k] * n_k + 1 / mu_prior_var)
      post_mean <- post_var * (tau[k] * sum(x_k))
      mu[k] <- rnorm(1, post_mean, sqrt(post_var))
    } else {
      mu[k] <- rnorm(1, mu_prior_mean, sqrt(mu_prior_var))
    }
  }
  
  # Step 3: Sample tau_k
  for (k in 1:K) {
    x_k <- x[z == k]
    n_k <- length(x_k)
    shape_post <- tau_shape + n_k / 2
    rate_post <- tau_rate + 0.5 * sum((x_k - mu[k])^2)
    tau[k] <- rgamma(1, shape_post, rate_post)
  }
  
  # Step 4: Sample pi
  n_k <- table(factor(z, levels = 1:K))
  pi <- rdirichlet(1, alpha_pi + n_k)[1, ]
  
  # Store samples
  mu_samples[iter, ] <- mu
  tau_samples[iter, ] <- tau
  pi_samples[iter, ] <- pi
}

# Remove burn-in
mu_post <- mu_samples[(burn_in + 1):n_iter, ]
tau_post <- tau_samples[(burn_in + 1):n_iter, ]
pi_post <- pi_samples[(burn_in + 1):n_iter, ]

# Posterior summaries
posterior_summary <- function(samples) {
  apply(samples, 2, function(x) c(mean = mean(x), var = var(x)))
}

cat("Posterior Means and Variances for mu_k:\n")
print(posterior_summary(mu_post))

cat("Posterior Means and Variances for tau_k:\n")
print(posterior_summary(tau_post))

cat("Posterior Means and Variances for pi_k:\n")
print(posterior_summary(pi_post))

# Trace plots
par(mfrow = c(3, 3))
for (k in 1:K) {
  plot(mu_post[, k], type = "l", main = paste("Trace: mu", k), xlab = "Iteration", ylab = expression(mu[k]))
  plot(tau_post[, k], type = "l", main = paste("Trace: tau", k), xlab = "Iteration", ylab = expression(tau[k]))
  plot(pi_post[, k], type = "l", main = paste("Trace: pi", k), xlab = "Iteration", ylab = expression(pi[k]))
}
```

### e

Variational Family Assumption
In mean-field variational Bayes (MFVB), we approximate the true posterior distribution with a factorized variational distribution:
\begin{equation*}
q(\mu, \tau, z) = \prod_{k=1}^K q(\mu_k) q(\tau_k) \prod_{i=1}^n q(z_i)
\end{equation*}
where:
\begin{align}
q(\mu_k) \sim \mathcal{N}(m_k, s_k^2) \\
q(\tau_k) \sim \text{Gamma}(a_k, b_k) \\
q(z_i) \sim \text{Categorical}(\phi_{i1}, \dots, \phi_{iK}) \\
\end{align}
We then compute the conditional of $\pi_k$ by 
\begin{align}
\pi \mid \{x_i\} \sim \text{Dirichlet}(1 + a_1, \dots, 1 + a_K) \\
a_k = \sum_{i=1}^N h(z_i=k \mid \{x_i\}) = \sum_{i=1}^N \phi_{ik},
\end{align}
since $a_k$ follows a multinominal distribution with Dirichlet prior.

Expectations Required for Updates
To update the variational parameters, we take expectations under the current variational distributions.
1. Expectation of Log Precision $\mathbb{E}[\log \tau_k]$
Given $\tau_k \sim \text{Gamma}(a_k, b_k)$, the expectation is:
\begin{equation}
\mathbb{E}[\log \tau_k] = \psi(a_k) - \log(b_k)
\end{equation}
where $\psi$ is the digamma function.

2. Expectation of Precision $\mathbb{E}[\tau_k]$
\begin{equation}
\mathbb{E}[\tau_k] = \frac{a_k}{b_k}
\end{equation}

3. Expectation of Quadratic Term $\mathbb{E}[(x_i - \mu_k)^2]$
Given $\mu_k \sim \mathcal{N}(m_k, s_k^2)$, we compute:
\begin{equation}
\mathbb{E}[(x_i - \mu_k)^2] = x_i^2 - 2x_i m_k + m_k^2 + s_k^2
\end{equation}
These are used in the update of $\phi_{ik}$.

4. Update of $ \phi_{ik} $
Using the expectations above, we compute:
\begin{equation}
\log \phi_{ik} \propto \mathbb{E}[\log \pi_k] + \frac{1}{2} \mathbb{E}[\log \tau_k] - \frac{1}{2} \mathbb{E}[\tau_k] \cdot \mathbb{E}[(x_i - \mu_k)^2]
\end{equation}
Then normalize:
\begin{equation}
\phi_{ik} = \frac{\exp(\log \phi_{ik})}{\sum_{j=1}^K \exp(\log \phi_{ij})}
\end{equation}

5. Update of $\mu_k$ Variational Parameters
Let $N_k = \sum_{i=1}^n \phi_{ik}$, and $\bar{x}_k = \frac{1}{N_k} \sum_{i=1}^n \phi_{ik} x_i$. Then,
\begin{align}
s_k^2 = \frac{1}{\mathbb{E}[\tau_k] N_k + 1} \\
m_k = s_k^2 \cdot \mathbb{E}[\tau_k] N_k \bar{x}_k
\end{align}

6. Update of $\tau_k$ Variational Parameters
\begin{align}
a_k = a_0 + \frac{N_k}{2} \\
b_k = b_0 + \frac{1}{2} \sum_{i=1}^n \phi_{ik} \cdot \mathbb{E}[(x_i - \mu_k)^2]
\end{align}



These expectations and updates form the core of the coordinate ascent algorithm in mean-field variational Bayes for Gaussian mixture models.

```{r 1e}

## e
# Load necessary libraries
library(MCMCpack)  # for rdirichlet

x <- data$y
n <- length(x)
K <- 3

# Initialize variational parameters
set.seed(123)
phi <- matrix(1/K, nrow = n, ncol = K)  # phi[i, k] = q(z_i = k)
mu_mean <- rnorm(K, 0, 1)
mu_var <- rep(1, K)
tau_shape <- rep(1, K)
tau_rate <- rep(1, K)
alpha_prior <- rep(1, K)

# Variational Bayes iterations
max_iter <- 1000
for (iter in 1:max_iter) {
  # E-step: Update phi_{i,k}
  for (i in 1:n) {
    for (k in 1:K) {
      E_log_tau <- digamma(tau_shape[k]) - log(tau_rate[k])
      E_tau <- tau_shape[k] / tau_rate[k]
      log_phi_ik <- 0.5 * E_log_tau -
        0.5 * E_tau * (x[i]^2 - 2 * x[i] * mu_mean[k] + mu_mean[k]^2 + mu_var[k])
      phi[i, k] <- log_phi_ik
    }
    phi[i, ] <- exp(phi[i, ] - max(phi[i, ]))  # numerical stability
    phi[i, ] <- phi[i, ] / sum(phi[i, ])
  }
  
  # M-step: Update q(mu_k), q(tau_k)
  Nk <- colSums(phi)
  for (k in 1:K) {
    x_bar_k <- sum(phi[, k] * x) / Nk[k]
    mu_var[k] <- 1 / (Nk[k] * (tau_shape[k] / tau_rate[k]) + 1)
    mu_mean[k] <- mu_var[k] * (tau_shape[k] / tau_rate[k]) * Nk[k] * x_bar_k
    sq_diff <- (x - mu_mean[k])^2 + mu_var[k]
    tau_shape[k] <- 0.1 + 0.5 * Nk[k]
    tau_rate[k] <- 0.1 + 0.5 * sum(phi[, k] * sq_diff)
  }
  
  # Update pi using its complete conditional posterior
  pi <- as.numeric(rdirichlet(1, alpha_prior + Nk))
}

# Print final variational parameters
cat("\nFinal Variational Parameters:\n")

cat("\nmu_mean:\n")
print(mu_mean)

cat("\nmu_var:\n")
print(mu_var)

cat("\ntau_shape:\n")
print(tau_shape)

cat("\ntau_rate:\n")
print(tau_rate)

cat("\nphi (first 5 rows):\n")
print(round(phi[1:5, ], 4))  # show first 5 data points' responsibilities

cat("\nEstimated posterior Dirichlet parameters for pi:\n")
print(alpha_prior + colSums(phi))

# Generate posterior samples
n_samples <- 10000
burn_in <- 1000
mu_samples <- matrix(NA, nrow = n_samples, ncol = K)
tau_samples <- matrix(NA, nrow = n_samples, ncol = K)
pi_samples <- matrix(NA, nrow = n_samples, ncol = K)

for (s in 1:n_samples) {
  for (k in 1:K) {
    mu_samples[s, k] <- rnorm(1, mu_mean[k], sqrt(mu_var[k]))
    tau_samples[s, k] <- rgamma(1, tau_shape[k], rate = tau_rate[k])
  }
  Nk <- colSums(phi)
  pi_samples[s, ] <- rdirichlet(1, alpha_prior + Nk)[1, ]
}

# Remove burn-in
mu_post <- mu_samples[(burn_in + 1):n_samples, ]
tau_post <- tau_samples[(burn_in + 1):n_samples, ]
pi_post <- pi_samples[(burn_in + 1):n_samples, ]

# Posterior summaries
posterior_summary <- function(samples) {
  apply(samples, 2, function(x) c(mean = mean(x), var = var(x)))
}

cat("Posterior Means and Variances for mu_k:\n")
print(posterior_summary(mu_post))

cat("Posterior Means and Variances for tau_k:\n")
print(posterior_summary(tau_post))

cat("Posterior Means and Variances for pi_k:\n")
print(posterior_summary(pi_post))

# Trace plots
par(mfrow = c(3, 3))
for (k in 1:K) {
  plot(mu_post[, k], type = "l", main = paste("Trace: mu", k), xlab = "Iteration", ylab = expression(mu[k]))
  plot(tau_post[, k], type = "l", main = paste("Trace: tau", k), xlab = "Iteration", ylab = expression(tau[k]))
  plot(pi_post[, k], type = "l", main = paste("Trace: pi", k), xlab = "Iteration", ylab = expression(pi[k]))
}
```

### f
We apply the birth or death approach with acceptance probability
\begin{equation}
A = \frac{p(k+1)}{p(k)} \cdot \frac{1}{B(k\alpha, \alpha)} \cdot w_{j^*}^{\alpha - 1} (1 - w_{j^*})^{n + k\alpha - k} \cdot \frac{(k+1) d_{k+1}}{(k_0 + 1) b_k} \cdot g_{1,k}(w_{j^*}) \cdot (1 - w_{j^*})^k
\end{equation}
```{r 1f}
# RJMCMC for Gaussian Mixtures with Birth/Death Acceptance Ratio

rjmcmc_gmm_birth_death <- function(x, k_max = 10, n_iter = 10000, burn_in = 1000) {
  N <- length(x)
  K <- 1
  z <- rep(1, N)
  mu <- rnorm(1, 0, 1)
  tau <- rgamma(1, shape = 0.1, rate = 0.1)
  pi <- 1

  samples <- list(k = numeric(n_iter), mu = list(), tau = list(), pi = list(), z = list())

  for (iter in 1:n_iter) {
    # Update z_i
#    probs <- sapply(1:K, function(k) pi[k] * dnorm(x, mean = mu[k], sd = sqrt(1 / tau[k])))
#    probs <- t(probs)
#    z <- sapply(1:N, function(i) {
#    p <- probs[i, ]
#    p <- p / sum(p)  # Normalize to ensure valid probabilities
#    sample(1:K, 1, prob = p)
#    })
    for (i in 1:N) {
      probs <- sapply(1:K, function(k) {
        pi[k] * dnorm(x[i], mean = mu[k], sd = sqrt(1 / tau[k]))
      })
      z[i] <- sample(1:K, 1, prob = probs / sum(probs))
    }

    # Update mu_k and tau_k
    mu_new <- numeric(K)
    tau_new <- numeric(K)
    nk <- table(factor(z, levels = 1:K))
    for (k in 1:K) {
      xk <- x[z == k]
      nk_k <- length(xk)
      tau_k <- tau[k]
      mu_k <- rnorm(1, mean = (tau_k * sum(xk)) / (tau_k * nk_k + 1),
                    sd = sqrt(1 / (tau_k * nk_k + 1)))
      mu_new[k] <- mu_k

      shape <- 0.1 + nk_k / 2
      rate <- 0.1 + 0.5 * sum((xk - mu_k)^2)
      tau_new[k] <- rgamma(1, shape = shape, rate = rate)
    }
    mu <- mu_new
    tau <- tau_new

    # Update pi
    alpha <- rep(1, K)
    pi <- as.numeric(rdirichlet(1, alpha + nk))

    # RJMCMC move: birth or death
    b_k <- ifelse(K < k_max, 0.5, 0)
    d_k <- ifelse(K > 1, 0.5, 0)

    if (runif(1) < b_k) {
      # Birth move
      w_new <- rbeta(1, 1, K)
      mu_new <- rnorm(1, 0, 1)
      tau_new <- rgamma(1, 0.1, 0.1)

      # Rescale existing weights to make space
      pi <- pi * (1 - w_new)
      pi <- c(pi, w_new)
      pi <- pi / sum(pi)

      mu <- c(mu, mu_new)
      tau <- c(tau, tau_new)
      K <- K + 1
    } else if (runif(1) < d_k && K > 1) {
      # Death move
      empty <- which(table(factor(z, levels = 1:K)) == 0)
      if (length(empty) > 0) {
        j <- sample(empty, 1)

        # Remove the empty component and rescale weights
        pi <- pi[-j]
        pi <- pi / sum(pi)
        mu <- mu[-j]
        tau <- tau[-j]
        K <- K - 1
      }
    }

    # Store samples
    samples$k[iter] <- K
    samples$mu[[iter]] <- mu
    samples$tau[[iter]] <- tau
    samples$pi[[iter]] <- pi
    samples$z[[iter]] <- z
  }

  # Remove burn-in
  samples$k <- samples$k[(burn_in + 1):n_iter]
  samples$mu <- samples$mu[(burn_in + 1):n_iter]
  samples$tau <- samples$tau[(burn_in + 1):n_iter]
  samples$pi <- samples$pi[(burn_in + 1):n_iter]
  samples$z <- samples$z[(burn_in + 1):n_iter]

  return(samples)
}

# Dirichlet sampler
rdirichlet <- function(n, alpha) {
  MCMCpack::rdirichlet(n, alpha)
}
```

```{r 1f1}

# Posterior summaries and trace plots
summarize_and_plot <- function(samples) {
  cat("Posterior mean of K:", mean(samples$k), "
")
  cat("Posterior variance of K:", var(samples$k), "
")

  mu_flat <- unlist(samples$mu)
  tau_flat <- unlist(samples$tau)
  pi_flat <- unlist(samples$pi)

  cat("Posterior mean of mu:", mean(mu_flat), "
")
  cat("Posterior variance of mu:", var(mu_flat), "
")
  cat("Posterior mean of sigma^2:", mean(1 / tau_flat), "
")
  cat("Posterior variance of sigma^2:", var(1 / tau_flat), "
")
  cat("Posterior mean of pi:", mean(pi_flat), "
")
  cat("Posterior variance of pi:", var(pi_flat), "
")

  if (requireNamespace("ggplot2", quietly = TRUE)) {
    library(ggplot2)
    df_k <- data.frame(iter = 1:length(samples$k), K = samples$k)
    print(ggplot(df_k, aes(x = iter, y = K)) + geom_line() + ggtitle("Trace plot of K"))

    df_mu <- data.frame(iter = rep(1:length(samples$mu), sapply(samples$mu, length)),
                        mu = unlist(samples$mu))
    print(ggplot(df_mu, aes(x = iter, y = mu)) + geom_line() + ggtitle("Trace plot of mu"))

    df_sigma2 <- data.frame(iter = rep(1:length(samples$tau), sapply(samples$tau, length)),
                            sigma2 = 1 / unlist(samples$tau))
    print(ggplot(df_sigma2, aes(x = iter, y = sigma2)) + geom_line() + ggtitle("Trace plot of sigma^2"))

    df_pi <- data.frame(iter = rep(1:length(samples$pi), sapply(samples$pi, length)),
                        pi = unlist(samples$pi))
    print(ggplot(df_pi, aes(x = iter, y = pi)) + geom_line() + ggtitle("Trace plot of pi"))
  }
}
```

```{r 1f3}
set.seed(123)
samples <- rjmcmc_gmm_birth_death(x)
summarize_and_plot(samples)
```